---
title: "Code_GSE142025"
author: "Pranto"
date: "`r Sys.Date()`"
output: html_document
---


### Import export counts matrix data
```{r}
file_list <- list.files(path = "GSE142025_RAW/", pattern = ".txt")

countsMatrix <- read.table(file = paste0("GSE142025_RAW/", file_list[1]), header = TRUE)
for (i in 2:length(file_list)) {
  tempCountMatrix <- read.table(file = paste0("GSE142025_RAW/", file_list[i]), header = TRUE)
  if(all(countsMatrix$Symbol==tempCountMatrix$Symbol))
  {
    countsMatrix <- cbind(countsMatrix, tempCountMatrix[2])
  }
  else{
    print("RowNames don't match...!")
  }
}
rownames(countsMatrix) <- countsMatrix[,'Symbol']
countsMatrix <- countsMatrix[,-1]
View(countsMatrix)

write.csv(countsMatrix, file = "CountsMatrix.csv")

# after the above steps we can now import our countsmatrix from CountsMatrix.csv file directly and rename the rowNames to first columns
countsMatrix <- read.csv("CountsMatrix.csv", header = TRUE, row.names = 1)
## Sort the count matrix according to the rownames
#countsMatrix <- countsMatrix[order(rownames(countsMatrix)), ]

#rownames(countsMatrix) <- countsMatrix[,'X']
#countsMatrix <- countsMatrix[, -1]
```

### Downloading the Ensemble Gene Id with biomaRt package
```{r}
#BiocManager::install("biomaRt")
library(biomaRt)
listEnsembl() #lists the available BioMart databases(Ensembl marts) hosted by Ensembl

#To know which datasets are available within a BioMart database, first select the BioMart database using useEnsembl and then use the listDatasets function on the selected Mart object.
selectedBioMartDBObject <- useEnsembl(biomart = "genes") #useEnsembl returns Ensembl marts type object

## list the available datasets in this Mart
datasets <- listDatasets(mart = selectedBioMartDBObject)
listMarts()
ensembl.connection <- useMart(biomart = 'ensembl', dataset = 'hsapiens_gene_ensembl')

attributes.list <- listAttributes(ensembl.connection)
filters.list <- listFilters(ensembl.connection)

num_of_row_to_check = nrow(countsMatrix)
countsMatrix <- head(countsMatrix, num_of_row_to_check)

## Get the ENSEMBL_GENE_ID for (num_of_row_to_check) genes
ensembl_gene_id <- getBM(attributes = c("ensembl_gene_id","external_gene_name"),
      filters = "external_gene_name",
      values = head(rownames(countsMatrix), num_of_row_to_check),
      mart = ensembl.connection)
```
## Mapping the Ensembl gene id with gene symbol
```{r}
## Sort the data frame according to the column
ensembl_gene_id <- ensembl_gene_id[order(ensembl_gene_id$external_gene_name), ]

head(ensembl_gene_id)
# it returns a logical array of duplicate gene name
duplicated(ensembl_gene_id$external_gene_name)
# It calculates number of duplicate row
sum(duplicated(ensembl_gene_id$external_gene_name))
# It shows the rownames of duplicated gene name
rownames(ensembl_gene_id[duplicated(ensembl_gene_id$external_gene_name), ])
## Getting genes without duplicate
ensembl_gene_id <-  ensembl_gene_id[!duplicated(ensembl_gene_id$external_gene_name), ]

rownames(head(countsMatrix, 10)) == ensembl_gene_id$external_gene_name

# Finding genes those don't have Gene_Ensembl_Id
pointer_to_count_matrix = 1
pointer_to_ensembl_gene_name = 1
counts_for_match_gene_id = 0

## This logical vector will indicate which genes have gene ensembl id. 
index_for_genes_those_have_gene_id <- rep(FALSE, num_of_row_to_check)

for (pointer_to_count_matrix in (1:num_of_row_to_check)) {
  if(pointer_to_ensembl_gene_name == (nrow(ensembl_gene_id) + 1))
  {
    break;
  }
  print(paste0("pcount = ", pointer_to_count_matrix," name = ", rownames(countsMatrix[pointer_to_count_matrix, ])));
  print(paste0("pensemble = ", pointer_to_ensembl_gene_name, " name = ", ensembl_gene_id[pointer_to_ensembl_gene_name, ]$external_gene_name));
  if(rownames(countsMatrix[pointer_to_count_matrix, ]) == ensembl_gene_id[pointer_to_ensembl_gene_name, ]$external_gene_name)
  {
    counts_for_match_gene_id = counts_for_match_gene_id + 1;
    pointer_to_ensembl_gene_name = pointer_to_ensembl_gene_name + 1;
    index_for_genes_those_have_gene_id[pointer_to_count_matrix] = TRUE;
    print("matched")
  }
  else
  {
    #print(pointer_to_count_matrix)
    #print(rownames(countsMatrix[pointer_to_count_matrix, ]))
    #print(pointer_to_ensembl_gene_name)
    #print(ensembl_gene_id[pointer_to_ensembl_gene_name, ]$external_gene_name)
  }
}
# Keeping only genes those have ensembl gene id
countsMatrix <- countsMatrix[index_for_genes_those_have_gene_id, ];
# checking the gene orders
all(rownames(countsMatrix) == ensembl_gene_id$external_gene_name);
# adding ensembl_gene_id to a seperate column in countsmatrix
countsMatrix$ensembl_gene_id = ensembl_gene_id$ensembl_gene_id;
# check if all the countmatrix rownames equal to the ensembe_gene_id's  external_gene_name column
all(rownames(countsMatrix) == ensembl_gene_id$external_gene_name)
# check if all the countmatrix's ensembl_gene_id column equal to ensembe_gene_id's  ensembl_gene_id column
all(ensembl_gene_id$ensembl_gene_id == countsMatrix$ensembl_gene_id)

# Export the new countmatrix with ensembl_gene_id with name CountsMatrix_With_Ensembl_Gene_id
write.csv(countsMatrix, file = "CountsMatrix_With_Ensembl_Gene_id.csv")
```

```{r}
#After store the CountsMatrix_With_Ensembl_Gene_id.csv I move the gene id column to first column so that's why I have to import and check it again if the order is same as `ensembl_gene_id$ensembl_gene_id`
countsMatrix_with_ensembl_gene_id <- read.csv("CountsMatrix_With_Ensembl_Gene_id.csv")
all(ensembl_gene_id$ensembl_gene_id == countsMatrix_with_ensembl_gene_id$ensembl_gene_id)
```


### Download colData and export to csv file
```{r}
library(GEOquery)
colData <- getGEO(GEO = "GSE142025", GSEMatrix = TRUE)
colData <- colData[["GSE142025_series_matrix.txt.gz"]]@phenoData@data

write.csv(colData, file = "ColData_Matrix_GSE142025.csv")
```
### Read the colData from the csv file and rename the rownames by first column
```{r}
colData <- read.csv(file = "ColData_Matrix_GSE142025.csv", row.names = 1)
#rownames(colData) <- colData[, 'X']
#colData <- colData[, -1]
```


## Check if all the columns name match in order with the title column of coldata matrix
```{r}
all(colnames(countsMatrix) == colData$title)
colnames(countsMatrix) <- rownames(colData)
```

### Here we rename the sample name to its group
```{r}
Control.ColDataIndex <- grepl("Control", colData$group.ch1)
Control.ColDataIndex
totalControlSample <- sum(Control.ColDataIndex)

Control.CountMatrix <- countsMatrix[, Control.ColDataIndex]
colnames(Control.CountMatrix) <- paste0("control", 1:totalControlSample)

colnames(countsMatrix)[Control.ColDataIndex] <- paste0("control", 1:totalControlSample)



Early_DN.ColDataIndex <- grepl("Early_DN", colData$group.ch1)
totalEarly_DNSample <- sum(Early_DN.ColDataIndex)
totalEarly_DNSample
rownames(colData)[Early_DN.ColDataIndex]
colnames(countsMatrix)[Early_DN.ColDataIndex] <- paste0("early", 1:totalEarly_DNSample)



Advanced_DN.ColDataIndex <- grepl("Advanced_DN", colData$group.ch1)
totalAdvanced_DNSample <- sum(Advanced_DN.ColDataIndex)
totalAdvanced_DNSample
rownames(colData)[Advanced_DN.ColDataIndex]
colnames(countsMatrix)[Advanced_DN.ColDataIndex] <- paste0("advanced", 1:totalAdvanced_DNSample)
```

```{r}
library(ggplot2)

ggplot(data = countsMatrix, aes(x = control1)) + geom_histogram(fill = "#525252", binwidth = 0.1) + labs(title = "control1")

histogram_plots <- list()

for (i in 1:totalControlSample) {
  sample_name <- paste0("control", i) #Generate the column name for sample
  sample_data <- countsMatrix[, sample_name] #Extract the sample data 
  
  #Creating a histogram for the current sample
  plot <- ggplot(data = data.frame(x = sample_data), aes(x = x)) +
    geom_histogram(fill = "#525252", binwidth = 0.1) +
    labs(title = sample_name)
  
  histogram_plots[[i]] <- plot
}
# Display the histogram plots
#for(i in 1:totalControlSample){
#  print(histogram_plots[[i]])
#}
library(gridExtra)

#Arrange the histogram plots in a 3x3 grid
grid.arrange(histogram_plots[[1]], histogram_plots[[2]], histogram_plots[[3]], histogram_plots[[4]], histogram_plots[[5]], histogram_plots[[6]], histogram_plots[[7]], histogram_plots[[8]], histogram_plots[[9]])
```
```{r}
countsMatrix_log2 <- log2(countsMatrix + 1)
ggplot(data = countsMatrix_log2, aes(x = control1)) + geom_histogram(fill = '#525252', binwidth = 0.01) + labs(title = "control1 of log2")
```


```{r}
library(reshape2)
df <- melt(countsMatrix, variable.name = "Samples")
df <- data.frame(df, condition = substr(df$Samples, 1, 5))

ggplot(df, aes(x = Samples, y = value, fill = condition)) + geom_boxplot() 
```


```{r}
ggplot(df, aes(x = value, color = Samples, fill = Samples)) + ylim(c(0, 0.25)) + geom_density(alpha = 0.2, size = 1.25)  + theme(legend.position = "right")
``` 
### MA - plot
```{r}
plot <- list()

#for(i in 23 : 27)
for(i in 2: 21)
{
  M = countsMatrix[, paste0('advanced',1)] - countsMatrix[, paste0('advanced', i)]
  A = (countsMatrix[, paste0('advanced', 1)] + countsMatrix[, paste0('advanced',i)]) / 2
  df = data.frame(A, M)
  plot[[i-1]] <- ggplot(df, aes(x = A, y = M)) + geom_point(size = 1.5, alpha = 1/5) + 
  geom_hline(yintercept = 0, color = "blue3") + stat_smooth(se = FALSE, method = "loess", color = "red3") + labs(title = paste0("advanced1 vs advanced", i))
}
#print(plot)
library(gridExtra)
grid.arrange(grobs = plot, ncol = 2)



### Ma plot for control samples
plot <- list()

#for(i in 23 : 27)
for(i in 2: 9)
{
  M = countsMatrix[, paste0('control',1)] - countsMatrix[, paste0('control', i)]
  A = (countsMatrix[, paste0('control', 1)] + countsMatrix[, paste0('control',i)]) / 2
  df = data.frame(A, M)
  plot[[i-1]] <- ggplot(df, aes(x = A, y = M)) + geom_point(size = 1.5, alpha = 1/5) + 
  geom_hline(yintercept = 0, color = "blue3") + stat_smooth(se = FALSE, method = "loess", color = "red3") + labs(title = paste0("control1 vs control", i))
}
#print(plot)
library(gridExtra)
grid.arrange(grobs = plot, ncol = 3)



### Ma plot for early_DN samples
plot <- list()

#for(i in 23 : 27)
for(i in 2: 6)
{
  M = countsMatrix[, paste0('early',1)] - countsMatrix[, paste0('early', i)]
  A = (countsMatrix[, paste0('early', 1)] + countsMatrix[, paste0('early',i)]) / 2
  df = data.frame(A, M)
  plot[[i-1]] <- ggplot(df, aes(x = A, y = M)) + geom_point(size = 1.5, alpha = 1/5) + 
  geom_hline(yintercept = 0, color = "blue3") + stat_smooth(se = FALSE, method = "loess", color = "red3") + labs(title = paste0("early1 vs early", i))
}
#print(plot)
library(gridExtra)
grid.arrange(grobs = plot, ncol = 3)
```


### Pheatmap
```{r}
mat.dist = countsMatrix
mat.dist = as.matrix(dist(t(mat.dist)))
mat.dist = mat.dist / max(mat.dist)
heatmap(mat.dist)
```



